An AI Success Story in Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument

### 2016-01-15 Papers/InProgress/IJCAI/paper.tex
paper update

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz \cite{ToDo:Leibniz}.  G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5}. The main difference to G\"odel's version is the
addition of a conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen 1998, p.365 \cite[p.365]{Hazen} states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight--it is
included in related manuscripts''. However, the omitted conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  that G\"odel's axioms were inconsistent.}

@@

For a computational analysis of G\"odel's ontological argument in this
approach the embedding of HOMLs such as K,
KB and S5 with various domain conditions (possibilist and actualist)
is required. This idea has been successfully followed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, translated to predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5}). The
resulting object logic is the HOML K with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. KB, S5) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
means of an existence predicate, which can be used to guard the
quantifiers. Therefore, the embedding approach is very flexible.

\subsection{Improved Embedding}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5}
\end{figure}

@@

The new embedding of \SFiveU in Isabelle is shown in Fig.~\ref{QML_S5}.
%
\marginpar{@Bruno: I commented you paragraph here, since I am not
  convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
  and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

Another evidence that the new embedding provides a significant performance boost 
is the successful automatic verification in Isabelle/HOL (with its automatic tactic Meson) of the modal
collapse \cite{Sobel}, which is one of the most strongly criticized
`side-effects' of G\"odel's and Scott's variants of the proof. In previous work \shortcite{C40}
the modal collapse has been proven by the higher-order provers
\textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
verification in the highly trusted Isabelle/HOL system still failed
\cite{J28}.  The success with the new embedding can be seen in Fig.~\ref{Scott_S5}.

\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/Scott_S5.png}}
\caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  Axioms;  Automatic Verification of Modal Collapse} \label{Scott_S5}
\end{figure}
%\marginpar{ToDo: this picture has to be replaced by one without Isabelle errors}

@@

In the embedding presented here, in Fig.~\ref{QML_S5}, advanced
syntax-sugaring effects provided by Isabelle were used to prevent
issues as enumerated above. The possibility to define boldface connectives allows us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{Scott_S5} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

@@

On the other hand, there is still room for improvements in
Isabelle/HOL. For example, Isabelle/HOL's Sledgehammer tool, which
orchestrates the calls to external automated theorem provers, such as
\textsc{Leo-II}, still fails to detect the inconsistency of G\"odel's
axioms, while a direct modeling of the problem in TPTP THF syntax in
combination with a direct call of \textsc{Leo-II} succeeds. In other
words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected. There are further
technical issues regarding Sledgehammer that motivate
improvements. For example, in default setting, Sledgehammer does not
immediately inform the user when a proof has been found and instead silently
 first executes a series of time-consuming proof analysis processes (e.g. it's
dependency minimizer), before it eventually informs the user of success. For the
involved proof problems presented here (e.g. G\"odel's theorem T3:
\textit{Necessarily, there exists God}), this phase of silence may
take several minutes (during which the user might actually give up on
the proof attempt), while \textsc{Leo-II} already reported success to
Sledgehammer much earlier (e.g. after 2,5 seconds for T3).

>>>>

St. Anselm's ontological argument \cite{Proslogion} can be regarded as
the ancestor of modern ontological arguments such as G\"odel's. In the
millenium between Anselm and G\"odel, many philosophers modified and
arguably improved Anselm's argument. Of particular importance to
G\"odel was the work of Leibniz \cite{ToDo:Leibniz}.  G\"odel's manuscript
(Fig.~\ref{GoedelScript}) can be considered a translation of Leibniz's
presentation of the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in Isabelle, is shown in
Fig.~\ref{Scott_S5U}. The main difference to G\"odel's version is the
addition of a conjunct in the definition of \emph{essence}. G\"odel's
different definition of essence can be seen either in his manuscript
(Fig.~\ref{GoedelScript}) or, in more modern notation, in the Isabelle
formalization shown in Fig.~\ref{InconsistencyIsabelleK}. For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required. This difference has been
considered inessential and merely an oversight by many. For instance,
Hazen 1998, p.365 \cite[p.365]{Hazen} states that ``G\"odel left this
clause out in [11], but this appears to have been an oversight--it is
included in related manuscripts''. However, the omitted conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_S5U}
where the model finder Nitpick \cite{Nitpick} confirms consistency).\footnote{In
  personal communication, Dana Scott confirmed that he was unaware
  that G\"odel's axioms were inconsistent.}

@@

For a computational analysis of G\"odel's ontological argument in this
approach the embedding of HOMLs such as K,
KB and S5 with various domain conditions (possibilist and actualist)
is required. This idea has been successfully followed in related work
\cite{C40}. The embedding of HOML is in fact
straightforward. Formulas in HOML are \emph{lifted}, translated to predicates
over worlds, which are themselves explicitly represented as
terms. The logical constants of HOML are translated to HOL terms in such a way that, for instance, 
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_o$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation
\cite{DBLP:journals/logcom/Ohlbach91}, which is here intra-logically
realized --- and extended for quantifiers --- in HOL by stating a set
of equations defining the logical constants (Fig.~\ref{QML_S5U}). The
resulting object logic is the HOML K with rigid terms and constant
domains (possibilist quantifiers). Other logics (e.g. KB, S5) can be
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
means of an existence predicate, which can be used to guard the
quantifiers. Therefore, the embedding approach is very flexible.

\subsection{Improved Embedding}\label{sec:improvedembedding}
\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/QML_S5U.png}}
\caption{Improved Embedding of \SFiveU} \label{QML_S5U}
\end{figure}

@@

The new embedding of \SFiveU in Isabelle is shown in
Fig.~\ref{QML_S5U}. It is important to note that we 
have  $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi$,
despite the differences in the underlying frame structures.

%
% \marginpar{@Bruno: I commented you paragraph here, since I am not
%   convinced (yet) that e.g. BF and CBF are equally valid in \SFiveU
%   and \SFive}
% \SFiveU is only slightly stronger than \SFive. Most importantly,
% $\vDash_{\SFive} \varphi$ iff $\vDash_{\SFiveU} \varphi^{U}$, where
% $\varphi^U$ is obtained from $\varphi$ by replacing all $\Box$ and
% $\Dia$ by, respectively, $\BlackBox$ and $\BlackDia$. Therefore,
% \SFiveU can be considered as adequate for metaphysics as \SFive
% (cf. \cite[p. 127]{WilliamsonModalLogicAsMetaphysics}). 
% % \cite{Mattey}, \cite[ToDo]{DunnHardegree}
%
With this improved embedding, the final theorem T3
(\textit{``Necessarily, there exists God''}) can be derived from
Scott's consistent version of the axioms fully automatically.  The fully
automatic proof has been generated (in about 2,5 seconds) by the theorem prover
\textsc{Leo-II}~\cite{C26} and subsequently verified in the proof
assistant Isabelle/HOL~\cite{NPW02}, as shown in Fig.~\ref{Scott_S5U}.
The collaboration between the two systems has been orchestrated by
Isabelle's Sledgehammer tool \cite{Sledgehammer}.

@@

Another evidence that the new embedding provides a significant performance boost 
is the successful automatic verification in Isabelle/HOL (with its automatic tactic Meson) of the modal
collapse \cite{Sobel}, which is one of the most strongly criticized
`side-effects' of G\"odel's and Scott's variants of the proof. In previous work \shortcite{C40}
the modal collapse has been proven by the higher-order provers
\textsc{Satallax} \cite{Satallax} and \textsc{Leo-II}, but a fully automatic
verification in the highly trusted Isabelle/HOL system still failed
\cite{J28}.  The success with the new embedding can be seen in Fig.~\ref{Scott_S5U}.

\begin{figure}[t]
\centerline{\includegraphics[width=\columnwidth]{./Images/Scott_S5U.png}}
\caption{Full Automation of T3 in \SFiveU; Consistency of Scott's
  Axioms;  Automatic Verification of Modal Collapse} \label{Scott_S5U}
\end{figure}
%\marginpar{ToDo: this picture has to be replaced by one without Isabelle errors}

@@

In the embedding presented here, in Fig.~\ref{QML_S5U}, advanced
syntax-sugaring effects provided by Isabelle were used to prevent
issues as enumerated above. The possibility to define boldface connectives allows us to drop the prefix; ``\texttt{binder}'' annotations enable modal quantifiers to be used in the standard binding way and reduce the need for explicit lambda abstractions; and a careful choice of priorities for infix connectives gives the parenthesis conventions that are more familiar to the user. As desired, the definition of essence in Fig.~\ref{Scott_S5U} is undeniably more immediately recognizable and comprehensible than the definition in Fig.~\ref{UglyEssence}. The embedding technique is now completely transparent to the user.

@@

On the other hand, there is still room for improvements. For example, Isabelle/HOL's Sledgehammer tool, which
orchestrates the calls to external automated theorem provers, such as
\textsc{Leo-II}, still fails to detect the inconsistency of G\"odel's
axioms, while a direct modeling of the problem in TPTP THF syntax in
combination with a direct call of \textsc{Leo-II} succeeds. In other
words, without independent experiments directly in TPTP THF the
inconsistency would thus not have been detected. There are further
technical issues regarding Sledgehammer that motivate
improvements. For example, in default setting, Sledgehammer does not
immediately inform the user when a proof has been found and instead silently
 first executes a series of time-consuming proof analysis processes (e.g. it's
dependency minimizer), before it eventually informs the user of success. For the
involved proof problems presented here (e.g. G\"odel's theorem T3:
\textit{Necessarily, there exists God}), this phase of silence may
take several minutes (during which the user might actually give up on
the proof attempt), while \textsc{Leo-II} already reported success to
Sledgehammer much earlier (e.g. after 2,5 seconds for T3).
